### 2017-07-06 inference.tex
Sec. 5 full.

The correspondence between the inference rules and the RG judgment of
Sec.~\ref{sec:reasoning} is stated thus\footnote{Proof can be found in
the supplementary material}:
\begin{theorem}
\label{thm:inference-sound}
  Forall i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then $\R \vdash \hoare{{\sf true}}{c}{\F}$
\end{theorem}
Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
\Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
\R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by the
inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule updated to use the inference judgment is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
  \R(\stl,\stg,\stg') \Leftrightarrow \I(\C{true},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \I(\C{false},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  \stable(\R_c,\F)\\
  \forall \stl,\stg.~ \F(\stl,\stg) \Rightarrow 
    G(\stg, \stl \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Since the inference judgment $c\elabsto \F$ is fully automatic,
automating the application of the \rulelabel{RG-Txn} rule for a
transaction now requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $\I$ are formulas in
the first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
encodes an $\SL$ expression in a FOL. Given a set expression $s$ of
$\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where $\phi$ encodes $s$ in
FOL, and $f$ is a boolean function that names the set defined by $s$.
Naming provides an easy handle to $s$, simplifying $\mssemof{\cdot}$.
Set variables ($x$, $\stl$, $\stg$, $\dots$) are encoded as
unconstrained boolean functions ($g_x$, $g_{\stl}$, $g_{\stg}$,
$\ldots$) that are named after the variable.  Encoding of a set
literal $\{a\,|\,\varphi\}$ introduces a new boolean function $g$ that
is $true$ for only those $a$ that satisfy $\varphi$.  Thus $g$
faithfully encodes the literal. The encoding of the $\existsl{}$
expression first eliminates the second-order quantification over sets.
Existentially bound set variables ($a$) is skolemized, i.e.,
instantiated with a new unbound variable $a'$, before encoding the
bound expression $s$ and the first-order constraint $\phi$. The
algorithm for encoding $\phi$, which is already a first-order formula,
is straightforward, hence not shown. It basically involves replacing
set variables with boolean functions named after the variables. The
first-order encoding for a bind expression describes the semantics of
set monad's $\bind$ in FOL. Let $s_1$ be a set, and let $f$ be a
function that maps each variable in $s_1$ to a new set. Then, $s_2 =
s_1 \bind f$ if and only if for all $y\in s_2$, there exists an $x \in
s_1$ such that $y = f(x)$, and forall $x\in s_1$, $f(x)\in s_2$. The
encoding adds new constraints to this effect. Conditional set
expressions and set union are encoded straightforwardly.

>>>>

The correspondence between the inference rules and the RG judgment of
Sec.~\ref{sec:reasoning} is stated thus\footnote{Proof can be found in
the supplementary material}:
\begin{theorem}
\label{thm:inference-sound}
  Forall i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{~c~}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}
Where the set expression $\F(\emptyset,\stg)$ has a natural
interpretation, on the lines discussed in the next sub-section.
% Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
% \Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
% \R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by the
inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule updated to use the inference judgment is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
  \R(\stl,\stg,\stg') \Leftrightarrow \I(\C{true},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \I(\C{false},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  \stable(\R_c,\F)\\
  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Since the inference judgment $c\elabsto \F$ is fully automatic,
automating the application of the \rulelabel{RG-Txn} rule for a
transaction now requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $\I$ are formulas in
the first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
encodes an $\SL$ expression in a FOL. Given a set expression $s$ of
$\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where $\phi$ encodes $s$ in
FOL, and $f$ is a boolean function that names the set defined by $s$.
Naming provides an easy handle to $s$, simplifying $\mssemof{\cdot}$.
Set variables ($x$, $\stl$, $\stg$, $\dots$) are encoded as
unconstrained boolean functions ($g_x$, $g_{\stl}$, $g_{\stg}$,
$\ldots$) that are named after the variable.  Encoding of a set
literal $\{a\,|\,\varphi\}$ introduces a new boolean function $g$ that
is $true$ for only those $a$ that satisfy $\varphi$.  Thus $g$
faithfully encodes the literal. The encoding of the $\existsl{}$
expression first eliminates the second-order quantification over sets.
Existentially bound set variable ($a$) is skolemized, i.e.,
instantiated with a new unbound variable $a'$, before encoding the
bound expression $s$ and the first-order constraint $\phi$. The
algorithm for encoding $\phi$, which is already a first-order formula,
is straightforward. It basically involves replacing set variables with
boolean functions named after the variables. The first-order encoding
for a bind expression describes the semantics of set monad's $\bind$
in FOL. Let $s_1$ be a set, and let $f$ be a function that maps each
variable in $s_1$ to a new set. Then, $s_2 = s_1 \bind f$ if and only
if for all $y\in s_2$, there exists an $x \in s_1$ such that $y =
f(x)$, and forall $x\in s_1$, $f(x)\in s_2$. The encoding adds new
constraints to this effect. Conditional set expressions and set union
are encoded straightforwardly.

Observe that the language of encoding shown in Fig.~\ref{fig:logic} is
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and boolean.
\end{itemize}
The fragment of FOL that satisfies the above syntactic conditions is
called Effectively Propositional first-order logic (denoted {\sf
EPR}), or Bernay-Sch{\"o}nfinkel-Ramsey (BSR) logic, which is known to
be decidable~\cite{z3epr}. The language of encoding, however, is a
combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
quantifier-free propositions ($\varphi$) that encode object language
expressions are drawn, and (b).  $\Prop^1$, the theory from which
invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
{\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}, hence:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
Isolation specifications presented in Sec.~\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$
fragment, making the verification a practical exercise.
